Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2023-08-19 analysis/paper/paper.Rmd
Minor edits

The potentially immense value of insights into population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a constant and direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @attenbrow2015; @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics, and other variables that might impact these [@attenbrow2015; @bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012]. In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD). We discuss how the data compare with each other as well as how they can potentially be drawn on to understand cultural historical developments in the region through the Mesolithic, c. 9500--4000 BCE.

# Background
Following the retreat of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979; @steffen2011]. The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or at least located close to the contemporaneous shoreline at their time of use. As coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating. Shoreline dating has a long tradition of use in Fennoscandian archaeology [@brogger1905; @hollender1901], and is especially useful for dating the large number of surveyed archaeological sites where temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available. Previous studies have suggested that shoreline dated sites can be used as a proxy to study temporal variation in population density and settlement patterns [@jorgensen2020; @mjærum2022; @solheim2018; @tallavaara2020].

Previously, the frequency of shoreline dated sites has been compared to RSPDs by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years [e.g. @solheim2018; @fossum2020; @mjærum2022; @tallavaara2020], or by the use of disjoint time-intervals of variable duration [@berg-hansen2022], which could cause issues when comparing the relative counts across these units [@bevan2021]. The point-based estimates do therefore not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between the frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the northeast to Arendal municipality in the southwest (see map in Figure \@ref(fig:map)). The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility. Thus, this geographical and temporal limit are also used for this study.

@@

Sites surveyed by means of test-pitting between Horten and Arendal were initially retrieved from the national heritage database Askeladden [@directorate2018], totalling at `r nrow(surveyed)` records. The records were then manually reviewed and given a quality score based on the criteria in Table \@ref(tab:tab1), indicating the degree to which the spatial location and extent of the sites is believed to be represented in the geometries available in the database [see also @roalkvam2020]. All sites with a quality score of 4 or worse were excluded from further analysis. Any sites situated at elevations that result in a shoreline date earlier than 9465 BCE were then excluded. This marks the latest start-date among the employed displacement curves, and no sites have yet been verified to hvae an older date than around 9300 BCE in the region [@glorstad2016]. Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2019. Only excavated sites with available spatial data in Askeladden or local databases at the Museum of Cultural History of the University of Oslo were included in the analysis. The 102 excavated sites in the dataset without relevant ^14^C-dates that were originally shoreline dated in the excavation reports were included in the SSPD along with the retained surveyed sites. This gave a total of 921 shoreline dated sites in the final SSPD.

@@

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in a version-controlled online repository at X. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017]. Analysis of the ^14^C-dates were done using the R packages *ADMUR* [@timpson2021] and *rcarbon* [@crema2021]. The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* for Monte Carlo simulation and model comparison.

@@

The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE search, using the differential evolution optimization algorithm *DEoptimR* [@brest2006]. The Monte Carlo simulations were then performed using the *SPDsimulationTest()* function from *ADMUR*. For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10000 times for each null model. The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the critical envelope is used to calculate a global p-value, indicating whether the null model can be rejected.

@@

```{r shoredate, echo = FALSE, fig.cap = "A) An example site (Askeladden ID 111967-1) relative to the isobases of the displacement curves.  B) The geologically derived shoreline displacement curves for the region and the curve interpolated to the example site. C) The black line on the y-axis represents the elevation of the site. This crosses the interpolated displacement curve at 7680 and 7490 BCE, which thus in practice gives an estimate of when the location of the site emerged from the sea and represents a \\textit{terminus post quem} date for the site. D) The shoreline dating approach used for this study, where the gamma distrbiution on the y-axis describes the likely vertical relationship between the shoreline and the site when it was in use.",  out.width = "500px", fig.align="center", regfloat = TRUE}

@@

When the frequency distribution of shoreline dates have been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of displacement curves, or by taking mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C). As outlined above,  these point-estimates are then typically aggregated in bins of 200 or 500 years. This approach is illustrated in Figure \@ref(fig:sumshore)B. The SSPD resulting from using the probabilistic method of shoreline dating, given in Figure \@ref(fig:sumshore)C, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is indicated by the shifts in the positions of peaks and troughs in the distribution after c. 8000 BCE.

@@

For the RSPD, the logistic model achieved a p-value of 0.072, and could therefore not be rejected (Figure \@ref(fig:spdmc)E). As is also indicated by the BIC values, of the compared models, the logistic model represents the best explanation of the data, while the failure to reject the model means that the data could be expected under this logistic development. This is characterised by a rapid growth from the start of the RSPD, around 8500 BCE, which plateaus at around 7500 BCE and then remains stable for the remainder of the period.

@@

Subsequently, the SSPD deviates negatively from the exponential null from c. 6800 to c. 5800 BCE. This is followed by a positive deviation in the last centuries before 5000 BCE. These developments should possibly been seen in relation to simultaneous developments in the inland areas of southern Norway. The negative deviation in the SSPD appears to correspond to an increase in inland activity, which is to take place in the period from c. 6700--5900 BCE and reach its peak around 6200--6000 BCE [@boaz1999; @selsing2010; @persson2018]. @boaz1999 related this to an economic reorganisation, involving a more focused exploitation of the interior and a reduction in the utilisation of coastal areas. This period is followed by a marked reduction in inland activity, lasting until c. 4500--4000 BCE [@persson2018, 204], the onset of which thus corresponds to the end of the negative deviation in the SSPD at c. 5800 BCE, and the positive deviation from c. 5300--5000 BCE. @boaz1999 associated the period of decrease in inland activity to the onset of the classic Nøstvet phase, which is now believed to commence in the middle of the 6th millennium BCE [@reitan2022]. The Nøstvet period is argued to be characterised by an increase in specialised coastal adaptation that also coincides with what @fuglestvedt2018[17] has been termed \'the great wave\' of rock art in Norway, starting from around 5700 BCE in south-eastern Norway [@fuglestvedt2018, 42--46; @glorstad2010, 216--233]. The onset of the classic Nøstvet thus roughly corresponds with the positive deviation in the SSPD around the later parts of the 6th millennium BCE.

Directly relating these patterns to demographic developments or land-use and mobility patterns is challenging given the data available here. Given the stationarity of the logistic model of the RSPD after c. 7500 BCE, it is possible that fluctuations in the SSPD reflect variation in mobility patterns that are unrelated to demographic developments. There appears to be a negative deviation from the logistic null model in the RSPD corresponding to the one in the SSPD just before 6000 BCE. However, the logistic model could not be rejected for the RSPD, and so it is inappropriate to interpret any deviations from the Monte Carlo envelope and place any analytical weight on the deviations in the RSPD, at least given the present data. While it is possible that the fairly limited sample of ^14^C-dates has led to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE [see also @solheim2018]. A related point that is also worth making here is that the overall demographic trends that the modelling efforts can hope to identify are likely to have been subject to smaller scale demographic fluctuations that simply cannot be ascertained given the available data and its resolution [cf. @timpson2021; @tallavaara2021], and which impacts the comparison of the two proxies following both from the resolution of the dating methods and from the different sample sizes underlying the SPDs.

Given the presence of people and corresponding signal in the SSPD before the start of the RSPD, and the suggested relevance of mobility patterns for the lacking signal in the RSPD in this earliest period, neither proxy is presumably devoid of influence from population density or mobility patterns. This is also given support by the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD in the first centuries of the 7th millennium BCE. However, properly assessing this relationship and the degree to which it has held constant throughout the period will depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.

Another relevant point to mention here is that the relative sea-level change is effectively built into the Monte Carlo simulations, much in the same way as the Monte Carlo simulation method used with radiocarbon dates account for fluctuations in the calibration curve [e.g. @crema2022]. The simulation envelopes therefore pertain to the frequency of shoreline dated sites that could be expected under the null models, given the distribution of sites in the study area (Figure \@ref(fig:incpolys)A), the corresponding local trajectories of relative sea-level change (Figure \@ref(fig:incpolys)B), and a constant relationship between sites and shoreline described by the gamma distribution employed for the dating procedure (Figure \@ref(fig:shoredate)D). Beyond uncertainties to do with the dating method itself, and by extension also the generation of the Monte Carlo envelope, it is also conceivable that variation in rates of sea-level change has impacted settlement patterns beyond that which could be directly expected under these null models. An increased movement instantiated by increased rates of relative sea-level change could result in the transition to a settlement pattern involving more frequent moves than that which could be expected under a constant adjustment to the shoreline. If this is the case, the proxy would still reflect degree of occupational duration, but this could be a response to the rate of relative sea-level change rather than other cultural and societal dimensions such as migratory events or adjustments to other environmental developments. Directly accounting for variation in the rate of sea-level change could thus prove important for properly assessing the substantive implications of variation in this proxy.

@@

The main results was that the RSPD is consistent with the logistic model fit to the data. This is characterised by a rapid logistic growth from the start of the RSPD, which plateaus and remains stationary from around 7500 BCE. This is contrasted to the shoreline dates, the development of which were better described by a function of exponential decay. While this indicates that some process of decline in the number of shoreline dated sites has occurred over time, the model was rejected as an adequate description of the SSPD. Some of the deviations from the model were then narratively explored to suggest some possible reasons for the reasons and nature of these deviations. Given the lack of a model that adequately describe the data, and the informal treatment of these deviations, these should be viewed as suggestions that represent potential lines for further interrogation.

While it appears that the SSPD and RSPD are differentially impacted by population density, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation. This is also relevant for the treatment of RSPDs, which in in Norwegian archaeology have usually been treated as constantly and uniformly impacted by population density through the period. We suggested that mobility patterns has relevance for the kind of pyrotechnology that was utilised in the first period of human occupation in the region, and that this is a determining contributor to the absence of radiocarbon dates from this earliest period. The degree and stability with which mobility patterns might have impacted the radiocarbon record from later periods remains unclear, and could prove important for isolating a demographic signal in the radiocarbon data -- a point that has also been made in other contexts [@hiscock2016, 219].

Some further lines of inquiry could be to draw on lithic technology and tool-kit diversity as a function of population size and density [e.g. @berg-hansen2018; @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g. @barton2014; @clark2017; @preston2018; @roalkvam2022], which is potentially also influenced by variable rates of relative sea-level change. In terms of mobility patterns, focus was here directed towards residential movement and occupational duration, but other dimensions relating for example to seasonal and territorial mobility could also prove to be important. In the same vein, environmental conditions, constraints and developments beyond relative sea-level change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g. @jorgensen2020; @ordonez2022; @lundstrom2023]. Furthermore, aligning the complex web of causes and effects that these different lines of evidence potentially represent will also be dependent on a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf. @timpson2021]. Finally, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data employed here, but could also provide important contextualisation for the observed development [see e.g. @lundstrom2021; @manninen2023; @solheim2018].

>>>>

The potentially immense value of insights into population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a constant and direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @attenbrow2015; @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics, and other variables that might impact these [@attenbrow2015; @crombe2014; @bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012]. In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD). We discuss how the data compare with each other as well as how they can potentially be drawn on to understand cultural historical developments in the region through the Mesolithic, c. 9500--4000 BCE.

# Background
Following the retreat of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979; @steffen2011]. The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or located close to the contemporaneous shoreline at their time of use. Thus, as coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating. Shoreline dating has a long tradition of use in Fennoscandian archaeology [@brogger1905; @hollender1901], and is especially useful for dating the large number of surveyed archaeological sites where temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available. A series of studies have suggested that shoreline dated sites can be used as a proxy to study temporal variation in population density [@jorgensen2020; @mjærum2022; @solheim2018; @tallavaara2020].

Previously, the frequency of shoreline dated sites has been evaluated by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years [e.g. @solheim2018; @fossum2020; @mjærum2022; @tallavaara2020], or by the use of disjoint time-intervals of variable duration [@berg-hansen2022], which could cause issues when comparing the relative counts across these units [@bevan2021]. However, the point-based estimates do not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between the frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the north-east to Arendal municipality in the south-west (see map in Figure \@ref(fig:map)). The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility. Thus, this geographical and temporal limit are also used for this study.

@@

Sites surveyed by means of test-pitting between Horten and Arendal were initially retrieved from the national heritage database Askeladden [@directorate2018], totalling at `r nrow(surveyed)` records. The records were then manually reviewed and given a quality score based on the criteria in Table \@ref(tab:tab1), indicating the degree to which the spatial location and extent of the sites is believed to be represented by the geometries available in the database [see also @roalkvam2020]. All sites with a quality score of 4 or worse were excluded from further analysis. Any sites situated at elevations that result in a shoreline date earlier than 9465 BCE were then excluded. This marks the latest start-date among the employed displacement curves, and no sites have yet been verified to have an older date than around 9300 BCE in the region [@glorstad2016]. Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2019. Only excavated sites with available spatial data in Askeladden or local databases at the Museum of Cultural History of the University of Oslo were included in the analysis. The 102 excavated sites in the dataset without relevant ^14^C-dates that were originally shoreline dated in the excavation reports were included in the SSPD along with the retained surveyed sites. This gave a total of 921 shoreline dated sites in the final SSPD.

@@

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in a version-controlled online repository at https://doi.org/10.17605/osf.io/fqwt5. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017]. Analysis of the ^14^C-dates were done using the R packages *ADMUR* [@timpson2021] and *rcarbon* [@crema2021]. The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* for Monte Carlo simulation and model comparison.

@@

The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE search, using the differential evolution optimization algorithm *DEoptimR* [@brest2006]. The Monte Carlo simulations were then performed using the *SPDsimulationTest()* function from *ADMUR*. For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10000 times for each null model. The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the simulations is used to calculate a global p-value, indicating whether the null model can be rejected.

@@

```{r shoredate, echo = FALSE, fig.cap = "A) The location of an example site (Askeladden ID 111967-1) relative to the isobases of the displacement curves.  B) The geologically derived shoreline displacement curves for the region and the curve interpolated to the example site. C) The black line on the y-axis represents the elevation of the site. This crosses the interpolated displacement curve at 7680 and 7490 BCE, which thus in practice gives an estimate of when the location of the site emerged from the sea and represents a \\textit{terminus post quem} date for the site. D) The shoreline dating approach used for this study, where the gamma function on the y-axis describes the likely vertical relationship between the shoreline and the site when it was in use.",  out.width = "500px", fig.align="center", regfloat = TRUE}

@@

When the frequency distribution of shoreline dates have been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of the displacement curves, or by taking mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C). As outlined above,  these point-estimates are then typically aggregated in bins of 200 or 500 years. This approach is illustrated in Figure \@ref(fig:sumshore)B. The SSPD resulting from using the probabilistic method of shoreline dating, given in Figure \@ref(fig:sumshore)C, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is indicated by the shifts in the positions of peaks and troughs in the distribution after c. 8000 BCE.

@@

For the RSPD, the logistic model achieved a p-value of 0.072, and could therefore not be rejected (Figure \@ref(fig:spdmc)E). As is also indicated by the BIC values of the compared models, the logistic model represents the best explanation of the data, and the failure to reject the model means that the data could be expected under this logistic development. This is characterised by a rapid growth from the start of the RSPD, around 8500 BCE, which plateaus at around 7500 BCE and then remains stable for the remainder of the period.

@@

Subsequently, the SSPD deviates negatively from the exponential null from c. 6800 to c. 5800 BCE. This is followed by a positive deviation in the last centuries before 5000 BCE. These developments should possibly been seen in relation to simultaneous developments in the inland areas of southern Norway. The negative deviation in the SSPD appears to correspond to an increase in inland activity, which is to take place in the period from c. 6700--5900 BCE and reach its peak around 6200--6000 BCE [@boaz1999; @selsing2010; @persson2018]. @boaz1999 related this to an economic reorganisation, involving a more focused exploitation of the interior and a reduction in the utilisation of coastal areas. This period is followed by a marked reduction in inland activity, lasting until c. 4500--4000 BCE [@persson2018, 204], the onset of which thus corresponds to the end of the negative deviation in the SSPD at c. 5800 BCE and the positive deviation from c. 5300--5000 BCE. @boaz1999 associated this period of decrease in inland activity to the onset of the classic Nøstvet phase, which is now believed to commence in the middle of the 6th millennium BCE [@reitan2022]. The Nøstvet period is argued to be characterised by an increase in specialised coastal adaptation that also coincides with what @fuglestvedt2018[17] has been termed \'the great wave\' of rock art in Norway, starting from around 5700 BCE in south-eastern Norway [@fuglestvedt2018, 42--46; @glorstad2010, 216--233]. The onset of the classic Nøstvet thus roughly corresponds with the positive deviation in the SSPD around the later parts of the 6th millennium BCE.

A recent study by @manninen2023, which considers a larger inland area of northern Scandinavia than the above-referenced studies found a corresponding increase in inland activity starting from the earliest parts of the Mesolithic that lasted until c. 6000 BCE. However, after this the population is found to have been characterised by a cessation of growth, rather than the clear decrease identified by the regional studies in southern Norway. Consequently, properly relating the developments in the coastal region to inland developments and their relevance to each other will benefit from an integrated approach that also explicitly considers the spatial dimension without invoking predefined analytical regions [cf. @crema2017].

Assessing the suggested and possibly varying influence of demographic developments and mobility patterns on the two employed proxies is challenging given the divergence between them without invoking additional variables that can . Following from the stationarity of the logistic model of the RSPD after c. 7500 BCE, it is possible that fluctuations in the SSPD reflect variation in mobility patterns that are unrelated to demographic developments. There does appear to be a negative deviation from the logistic null model in the RSPD corresponding to the one in the SSPD just before 6000 BCE. However, the logistic model could not be rejected for the RSPD, and so it is inappropriate to interpret any deviations from the Monte Carlo envelope and place any analytical weight on the deviations in the RSPD, at least given the present data. While it is possible that the fairly limited sample of ^14^C-dates has led to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE [see also @solheim2018]. A related point that is also worth making here is that the overall demographic trends that the modelling efforts can hope to identify are likely to have been subject to smaller scale demographic fluctuations that simply cannot be ascertained given the available data and its resolution [cf. @tallavaara2021], and which impacts the comparison of the two proxies following both from the different resolutions of the dating methods and the sample sizes underlying the SPDs.

Given the presence of people and corresponding signal in the SSPD before the start of the RSPD, and the suggested relevance of mobility patterns for the lacking signal in the RSPD in this earliest period, neither proxy is presumably devoid of influence from population density or mobility patterns. This is also given support by the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD in the first centuries of the 8th millennium BCE. However, properly assessing this relationship and the degree to which it has held constant throughout the period will depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.

Another relevant point to mention here is that the relative sea-level change is effectively built into the Monte Carlo simulations, much in the same way as the Monte Carlo simulation method used with radiocarbon dates account for fluctuations in the calibration curve [e.g. @crema2022]. The simulation envelopes therefore pertain to the frequency of shoreline dated sites that could be expected under the null models, given the distribution of sites in the study area (Figure \@ref(fig:incpolys)A), the corresponding local trajectories of relative sea-level change (Figure \@ref(fig:incpolys)B), and a constant relationship between sites and shoreline described by the gamma function employed for the dating procedure (Figure \@ref(fig:shoredate)D). Beyond uncertainties to do with the dating method itself, and by extension also the generation of the Monte Carlo envelope, it is also conceivable that variation in rates of sea-level change has impacted settlement patterns beyond that which could be directly expected under these null models. An increased movement instantiated by increased rates of relative sea-level change could result in the transition to a settlement pattern involving more frequent moves than that which could be expected under a constant adjustment to the shoreline. If this is the case, the proxy would still reflect degree of occupational duration, but this could be a response to the rate of relative sea-level change rather than other cultural and societal dimensions such as migratory events or adjustments to other environmental developments. Directly accounting for variation in the rate of sea-level change could thus prove important for properly assessing the substantive implications of the fluctuations in this proxy.

@@

The main results was that the RSPD is consistent with the logistic model fit to the data. This is characterised by a rapid logistic growth from the start of the RSPD, which plateaus and remains stationary from around 7500 BCE. This is contrasted to the shoreline dates, the development of which were better described by a function of exponential decay. While this indicates that some process of decline in the number of shoreline dated sites has occurred over time, the model was rejected as an adequate description of the SSPD. Some of the deviations from the model were then narratively explored to suggest some possible reasons for these deviations. However, given the lack of a model that adequately describe the data, and the informal treatment of these deviations, these should be viewed as suggestions that represent potential lines for further interrogation.

While it appears that the SSPD and RSPD are differentially impacted by population density, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation. This is also relevant for the treatment of RSPDs in the region, as we suggested that mobility patterns has relevance for the kind of pyrotechnology that was utilised in the first period of human occupation in the region, and that this is a contributor to the absence of radiocarbon dates from this earliest period. The degree and stability with which mobility patterns might have impacted the radiocarbon record from later periods, potentially as mediated by pyrotechnology, remains unclear, and could prove important for properly isolating a demographic signal in the radiocarbon data -- a point that has also been made in other contexts [@crombe2014; @hiscock2016, 219].

Some further lines of inquiry could be to draw on lithic technology and tool-kit diversity as a function of population size and density [e.g. @berg-hansen2018; @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g. @barton2014; @clark2017; @preston2018; @roalkvam2022], which is potentially also influenced by variable rates of relative sea-level change. In terms of mobility patterns, focus was here directed towards residential movement and occupational duration, but other dimensions relating for example to seasonal and territorial mobility, as reflected in the discussion of inland activity, could also prove to be important. In the same vein, environmental conditions, constraints and developments beyond relative sea-level change also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g. @jorgensen2020; @ordonez2022; @lundstrom2023]. Furthermore, aligning the complex web of causes and effects that these different lines of evidence potentially represent will also benefit from a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf. @timpson2021]. Finally, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data employed here, but could also provide important contextualisation for the observed development [see e.g. @lundstrom2021; @manninen2023; @solheim2018].
